---
title: "SASP enpoint survival"
author: "Daniel Schwartz"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
  - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---


# Set working environment; load packages

```{r setup, message = FALSE, warning = FALSE}
# Clear and set working directory 
rm(list = ls())
setwd("~/GitHub/ssp_phage/data")
library("dplyr")
library("tidyr")
library("readr")
library("broom") #to extract stat results as table
# d <- read.csv("~/GitHub/ssp_phage/data/HeatDecayperc.survivals.csv", header = TRUE)
library(here)
```

# Load data

```{r}
d <- read_csv(here("data","ssp_data.csv"))%>%
  # to avoid confusion with R commands change group to g
  mutate(group=gsub("group.", "g", group))%>%
	
  #remove data from prelims
  filter(group!="preliminary")%>%
	
  #remove data on filtered samples
  filter(!filter)%>%
	
  #Calculate titer from counts
  mutate(counts=as.numeric(counts ))%>%
  mutate(no.ml = counts/(dilution*ml.plated))%>%
	
  #calculate percent survival relative to T=0 (no heat treatment)
  group_by(host,phage,assay, group)%>%
  mutate(perc.survival = 100*no.ml/no.ml[heat.min==0])%>%
  
  # keep endpoint values
  filter(heat.min==45)%>%
  
  #add spore type 
  mutate(spore=case_when(assay=="PFU" ~ "viral spore",
                           phage=="noPhage" ~ "non-infected host spore",
                           TRUE ~ "infected host spore"))%>%

  # make a colony variable
  mutate(colony=interaction(host,group))%>%
  mutate(colony=paste0(host,"-col",group))%>%
  mutate(colony=gsub("colg","col",colony))%>%
  ungroup()

  #export data
  write_csv(d,here("data/percent_T-end.csv"))
```

# Does deletion of sspAB diminish perc.survival?

Bacillus culture was split into three phage treatments: (SPO1, Goe2, noPhage).
The "group" column refers to spores that were assayed on same day. 
The "colony" column reflects pairing to be accounted for in analysis. 

```{r}
# subset perc.survivals by colony origin
cfu.1 <- filter(d, assay == "CFU" & phage == "noPhage") %>%
  dplyr::select(host, assay, colony, perc.survival)

# extract perc.survivals from wildtype
x.wt <-  cfu.1 %>%
  filter(host == "wt") %>%
  pull(perc.survival)

# extract perc.survivals from mutant
y.mut <-  cfu.1 %>%
  filter(host == "dsspAB") %>%
  pull(perc.survival)

# conduct one-sided, paired t-test
cfu.1.test1 <- t.test(x.wt, y.mut, alternative = "greater")

# conduct two-sided, paired t-test
cfu.1.test2 <- t.test(x.wt, y.mut, alternative = "two.sided")

# Answer: deletion of sspAB reduces heat tolerance
# two-sided: P = 0.003983 (alternative = "two.sided")
# one-sided: P = 0.001992 (alternative = "greater")
```


# Does infection with phage ameliorate effects of sspAB deletion?

For each colony we calculate the ratio of survivng spores (%) in phage infected vs non-infected cultures. The null hypthesis is that phage has no effect, so that the ratio eqals 1. The alternative is that a greater percentage of spores survive in phage infected cultures, so we do a one sided test.

```{r}
# subset perc.survivals by host, phage, and group
cfu.2 <- filter(d, assay == "CFU", host == "dsspAB") %>%
  dplyr::select(host, phage, assay, colony, perc.survival)  

# convert to wide
cfu.2 <- cfu.2 %>% 
  # spread(phage, perc.survival)
  tidyr::pivot_wider(names_from = phage, values_from = perc.survival)

# calculate ratio of perc.survivals between no-phage and plus-phage treatments
cfu.2 <- mutate(cfu.2, ratio.spo1 = cfu.2$SPO1/cfu.2$noPhage,
                       ratio.goe2 = cfu.2$Goe2/cfu.2$noPhage)

# Tests

# 1) Goe2: one-sided, paired t-test
cfu.Goe2.test <- t.test(cfu.2$ratio.goe2, mu = 1, alternative = "greater")
# one-sided: P = 0.1268 (alternative = "greater" -> does phage ameliorate?)
# answer = no

# 2) SPO1: one-sided, paired t-test
cfu.SPO1.test <- t.test(cfu.2$ratio.spo1, mu = 1, alternative = "greater")
# one-sided: P = 0.0792
# answer = maybe

# 3) Phage (Goe2 and SPO1), paired t-test: does "phage" ameliorate
cfu.both.test <- t.test(c(cfu.2$ratio.goe2, cfu.2$ratio.spo1), mu = 1, alternative = "greater")
# one-sided: P = 0.01634
# answer = yes

# 4) Ratio in Goe2 vs SPO1 perc.survivals, paired
cfu.GvsS.test <- t.test(x=cfu.2$Goe2, y=cfu.2$SPO1, mu = 0, alternative = "two.sided")
# two-sided: P = 0.0006878
# answer = yes (SPO1 ameliorates more than Goe2)

cfu.table <- dplyr::bind_rows(
  broom::glance(cfu.Goe2.test),
  broom::glance(cfu.SPO1.test),
  broom::glance(cfu.both.test),
  broom::glance(cfu.GvsS.test)
)
```

#  Does infection with phage ameliorate effects of sspAB deletion in virospores?   
Compare perc.survivals of viral spores to perc.survivals of colony-forming spores in the WT. 
There are no viral spores in non-infected cultures. 

```{r}
# subset perc.survivals by deletion host, phage treatment, and group
pfu <- filter(d,  host == "dsspAB") %>%
  # select CFU for noPhage and PFU for infected cultures
  filter((phage == "noPhage" &  assay == "CFU") | (assay == "PFU")) %>%
  dplyr::select(host, phage,  colony, perc.survival)  

# convert to wide
pfu <- pfu %>% 
  # spread(phage, perc.survival)
  tidyr::pivot_wider(names_from = phage, values_from = perc.survival)

# calculate differences in perc.survivals between no-phage and plus-phage treatments
pfu <- mutate(pfu, ratio.spo1 = pfu$SPO1/pfu$noPhage ,
                   ratio.goe2 = pfu$Goe2/pfu$noPhage)

# Tests

# 1) Goe2: one-sided, paired t-test
pfu.Goe2.test <- t.test(pfu$ratio.goe2, mu = 1, alternative = "greater")
# one-sided: P = 0.1521
# answer: no

# 2) SPO1: one-sided, paired t-test
pfu.SPO1.test <- t.test(pfu$ratio.spo1, mu = 1, alternative = "greater")
# one-sided: P = 0.1802
# answer: no

# 3) Phage (Goe2 and SPO1)
pfu.both.test <- t.test(c(pfu$ratio.goe2, pfu$ratio.spo1), mu = 1, 
                           alternative = "greater")
# one-sided: P = 0.05531
# answer: maybe

# 4) Difference in Goe2 vs SPO1 
pfu.GvsS.test <- t.test(x=pfu$Goe2, y=pfu$SPO1, mu = 0, alternative = "two.sided")
# two-sided: 0.8137
# answer: no

pfu.table <-dplyr::bind_rows(
  broom::glance(pfu.Goe2.test),
  broom::glance(pfu.SPO1.test),
  broom::glance(pfu.both.test),
  broom::glance(pfu.GvsS.test)
)
```


#  Do colony-forming spores and viral spores differ from each other when infecting the mutant host?  
One-sided hypothesis: viral spores are more resistant. 
That is, they have a greater perc.survival value: perc.survival.PFU/perc.survivalPFU > 1

```{r}
# subset perc.survivals by deletion host, assay type (CFU vs. PFU), and colony
cfu.pfu <- filter(d,  host == "dsspAB", phage != "noPhage") %>%
    dplyr::select(host, phage, assay , colony, perc.survival)  

# convert to wide
cfu.pfu  <- cfu.pfu  %>% 
  # spread(phage, perc.survival)
  tidyr::pivot_wider(names_from = assay, values_from = perc.survival)

# calculate differences in perc.survivals between CFU and PFU
cfu.pfu  <- mutate(cfu.pfu , dif = PFU/CFU)

cfu.pfu.Goe2.test <-
  cfu.pfu %>%
  filter(phage  =="Goe2") %>%
  pull(dif) %>%
  t.test(., mu = 1, alternative = "greater")
# one-sided: P = 0.507

cfu.pfu.SPO1.test <-
  cfu.pfu %>%
  filter(phage == "SPO1") %>%
  pull(dif) %>%
  t.test(., mu = 1, alternative = "greater")
# one-sided: P = 0.958
# answer: no

cfu.pfu.phage.test <-  
  cfu.pfu %>%
  t.test(dif~phage, mu = 0, alternative = "two.sided", data =.)
# one-sided: P = 0.427
# answer: no

cfu.pfu.table <- dplyr::bind_rows(
  broom::glance(cfu.pfu.Goe2.test),
  broom::glance(cfu.pfu.SPO1.test),
  broom::glance(cfu.pfu.phage.test)
)
```

# Does infection with phage compromise resistance of WT spores?
```{r}
cfu.3 <- filter(d, assay == "CFU", host == "wt") %>%
  dplyr::select(host, phage, assay, colony, perc.survival)  

# convert to wide
cfu.3 <- cfu.3 %>% 
  # spread(phage, perc.survival)
  tidyr::pivot_wider(names_from=phage, values_from=perc.survival)

# calculate difference and test
cfu.3 <- mutate(cfu.3, ratio.spo1 = cfu.3$SPO1/cfu.3$noPhage,
                       ratio.goe2 = cfu.3$Goe2/cfu.3$noPhage)

cfu.3.Goe2.test <- t.test(cfu.3$ratio.goe2, mu = 1, alternative = "less")
# one-sided: P = 0.0794

cfu.3.SPO1.test <- t.test(cfu.3$ratio.spo1, mu = 1, alternative = "less")
# one-sided: P = 0.374

cfu.3.phage.test <- t.test(c(cfu.3$ratio.goe2, cfu.3$ratio.spo1), mu = 0, 
                           alternative = "two.sided")
# one-sided: P =  0.00898

cfu.wt.table <- dplyr::bind_rows(
  broom::glance(cfu.3.Goe2.test),
  broom::glance(cfu.3.SPO1.test),
  broom::glance(cfu.3.phage.test)
)
```